---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# harpINLA

**harpINLA** is an example repo for fitting a spatio-temporal log Gaussian Cox Process model to animal telemetry data using R-INLA and follows the "going off grid" approach outlined in Simpson et al. 2016 https://doi.org/10.1093/biomet/asv064.

To address issues with approximation across land boundaries, the spatial mesh is defined using the Bakka Barrier model: https://doi.org/10.1016/j.spasta.2019.01.002

All the data objects required for the analysis are in the `/data` folder

## Example code to run analysis

### Load libraries and import data

```{r lgcp example set up, message = F, warning = F}
# Load libraries
require(INLA)
require(inlabru)
require(tidyverse)
require(sf)
require(raster)
require(rgeos)

# Load data from repo
dat <- readRDS("data/harps2500_indexed.rds") # random subsample of animal locations
land <- readRDS("data/land_sf.rds") # land shapefile - mainly for plotting
bound <- readRDS("data/boundary.rds") # boundary of the region for the INLA barrier model
mesh <- readRDS("data/mesh.rds") # INLA mesh
dmesh <- readRDS("data/dmesh.rds") # dmesh containing the Voroni polygons
mesh_ice <- readRDS("data/mesh_ice.rds") # environmental covariate values for the dmesh

# Define an Albers equal area projection using kilometres as units
prj = "+proj=laea +lat_0=75 +lon_0=-25 +x_0=0 +y_0=0 +units=km +no_defs +ellps=WGS84"

```


### Define the barrier model

```{r barrier model, message = F, warning = F}
# Set up boundary matern model
tl = length(mesh$graph$tv[,1]) # the number of triangles in the mesh

# Compute triangle positions
posTri = matrix(0, tl, 2) 
for (t in 1:tl){
  temp = mesh$loc[mesh$graph$tv[t, ], ]
  posTri[t,] = colMeans(temp)[c(1,2)] 
}
posTri = SpatialPoints(posTri, proj4string = CRS(prj))

normal = unlist(over(bound, posTri, returnList = T)) # check which mesh triangles are inside the normal area
barrier.triangles = setdiff(1:tl, normal)
poly.barrier = inla.barrier.polygon(mesh, barrier.triangles)

# create the matern object
barrier.model = inla.barrier.pcmatern(mesh,
                                      barrier.triangles = barrier.triangles,
                                      prior.range = c(50, .1),
                                      prior.sigma = c(10, 0.01))
```


### Define the time mesh for the annual migration

```{r time mesh, message = F, warning = F}
# Create a 1d time mesh for the annual cycle
tmesh <- inla.mesh.1d(loc = 1:4, boundary = "free") # boundary needs to be changed to 'cyclic' but this increases computation time
k <- length(tmesh$loc) # number of time periods

# number of time groups (4 seasons)
idx <- inla.spde.make.index(name = 's',
                            n.spde = barrier.model$f$n,
                            n.group = 4)

```


### Define exposure for the LGCP model
```{r define exposure, message = F, warning = F}
# Calculate weights for each polygon in the dmesh based on polygon area
# Set weights to 0 when polygon is outside study area

w <- sapply(1:length(dmesh), function(i){
  if(gIntersects(dmesh[i,], bound))
    return(gArea(gIntersection(dmesh[i,], bound)))
  else return(0)
})

# Exposure for LGCP is the weights of the dmesh replicated by the number of seasons in the time mesh
st.vol <- rep(w, k)
```


### Format space-time sea ice covariate

Distribution of animals is in part driven by sea ice concentration. This can be included as a smooth term in the model.

However, the location data come from different time periods (1990s, 2000s, 2010s) - with different sea ice conditions.

The seasonal sea ice data for each season and each time period associated with each Voroni polygon is contained in `mesh_ice`.

Format ice data into a vector ready to add to a INLA stack for each time period.

```{r sea ice, message = F, warning = F}
# format for the stack will be the covariate at each polygon for each time point in the 1d mesh
# append the covariate for the location to the end of the vector
# repeat for each stack - labelled 1,2,3,5 as we have no location data for time period 4
ice_1 <- c(mesh_ice[,1], mesh_ice[,2], mesh_ice[,3], mesh_ice[,4], dat$ice[dat$year_i == 1]) # seasonal ice and add location ice
ice_2 <- c(mesh_ice[,5], mesh_ice[,6], mesh_ice[,7], mesh_ice[,8], dat$ice[dat$year_i == 2])
ice_3 <- c(mesh_ice[,9], mesh_ice[,10], mesh_ice[,11], mesh_ice[,12], dat$ice[dat$year_i == 3])
ice_5 <- c(mesh_ice[,17], mesh_ice[,18], mesh_ice[,19], mesh_ice[,20], dat$ice[dat$year_i == 5])
```

### Create INLA stacks
```{r inla stack, message = F, warning = F}
# Stack 1
# Create A matrix with space and time indexing
Ast_1 <- inla.spde.make.A(mesh = mesh,
                          loc = cbind(dat$x[dat$year_i == 1], dat$y[dat$year_i == 1]),
                          group = dat$season[dat$year_i == 1],
                          n.group = k,
                          group.mesh = tmesh)

n <- nrow(dat[dat$year_i == 1,]) # number of locations
m <- barrier.model$f$n # number of mesh nodes

y <- rep(0:1, c(k * m, n))
expected <- c(st.vol, rep(0, n)) # this is specific to the number of locations in each time point

stk_1 <- inla.stack(
  data = list(y = y,
              expect = expected), 
  A = list(rbind(Diagonal(n = k * m), Ast_1), 1), 
  effects = list(idx,
                 list(a0 = rep(1, (k * m) + n),
                      ice = ice_1)))

# Stack 2
# Create A matrix with space and time indexing
Ast_2 <- inla.spde.make.A(mesh = mesh,
                          loc = cbind(dat$x[dat$year_i == 2], dat$y[dat$year_i == 2]),
                          group = dat$season[dat$year_i == 2],
                          n.group = k,
                          group.mesh = tmesh)

n <- nrow(dat[dat$year_i == 2,]) # number of locations
m <- barrier.model$f$n # number of mesh nodes

y <- rep(0:1, c(k * m, n))
expected <- c(st.vol, rep(0, n))

stk_2 <- inla.stack(
  data = list(y = y,
              expect = expected), 
  A = list(rbind(Diagonal(n = k * m), Ast_2), 1), 
  effects = list(idx,
                 list(a0 = rep(1, (k * m) + n),
                      ice = ice_2)))

# Stack 3
# Create A matrix with space and time indexing
Ast_3 <- inla.spde.make.A(mesh = mesh,
                          loc = cbind(dat$x[dat$year_i == 3], dat$y[dat$year_i == 3]),
                          group = dat$season[dat$year_i == 3],
                          n.group = k,
                          group.mesh = tmesh)

# Create lgcp stack
n <- nrow(dat[dat$year_i == 3,]) # number of locations
m <- barrier.model$f$n # number of mesh nodes

y <- rep(0:1, c(k * m, n))
expected <- c(st.vol, rep(0, n))

stk_3 <- inla.stack(
  data = list(y = y,
              expect = expected), 
  A = list(rbind(Diagonal(n = k * m), Ast_3), 1), 
  effects = list(idx,
                 list(a0 = rep(1, (k * m) + n),
                      ice = ice_3)))

# Stack 5
# Create A matrix with space and time indexing
Ast_5 <- inla.spde.make.A(mesh = mesh,
                          loc = cbind(dat$x[dat$year_i == 5], dat$y[dat$year_i == 5]),
                          group = dat$season[dat$year_i == 5],
                          n.group = k,
                          group.mesh = tmesh)

# Create lgcp stack
n <- nrow(dat[dat$year_i == 5,]) # number of locations
m <- barrier.model$f$n # number of mesh nodes

y <- rep(0:1, c(k * m, n))
expected <- c(st.vol, rep(0, n))

stk_5 <- inla.stack(
  data = list(y = y,
              expect = expected), 
  A = list(rbind(Diagonal(n = k * m), Ast_5), 1), 
  effects = list(idx,
                 list(a0 = rep(1, (k * m) + n),
                      ice = ice_5)))

# join stacks together
stk <- inla.stack(stk_1, stk_2, stk_3, stk_5)
```

### Prior specification and model formula
Due to the large number of unique covariate values, use the `inla.group` function.

```{r prior, message = F, warning = F}
# PC prior on time mesh
pcrho <- list(prior = 'pccor1', param = c(0.7, 0.7))

# formula for lgcp with ice covariate and barrier model spde
form <- y ~ 0 + a0 + 
  f(inla.group(ice,
               n = 25,
               method = "cut"),
    model = "rw2",
    scale.model = T,
    hyper = list(theta = list(prior="pc.prec",
                              param=c(1, 0.01)))) + 
  f(s,
    model = barrier.model,
    group = s.group,
    control.group = list(model = 'ar1',
                         hyper = list(theta = pcrho)))

```

### Fit the model
NB. This takes around 2.5 hours on a 2.7 GHz Intel i5 with 16GB RAM

```{r fitting, message = F, warning = F, eval = F}
# fit the model
fit <- inla(form,
            family = 'poisson',
            data = inla.stack.data(stk),
            E = inla.stack.data(stk)$e,
            control.predictor = list(A = inla.stack.A(stk)),
            control.inla = list(strategy = 'adaptive'),
            control.compute = list(dic = T,
                                   waic = T,
                                   config = T),
            verbose = T)
```

